Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support).

### 2017-02-09 csf-2017/paper.tex
Some more discussion of the capability machine.

\begin{figure*}
  \centering
  \begin{equation*}
    \Phi  \rightarrow
    \begin{cases}
      \sem{\decode(\memheap(\addr))}(\Phi) & \text{if $\memreg(\pcreg) = \stdcap$ and $\start \leq \addr \leq \addrend$ and $\perm \in \{ \exec,\rwx, \rwlx \}$ } \\
      \failed                                 & \text{otherwise}
    \end{cases}
  \end{equation*}%
  \begin{align*}
    \updatePcPerm{w} &=
                       \begin{cases}
                         ((\exec,\gl),\start,\addrend,\addr) & \text{if $w = ((\entry,\gl),\start,\addrend,\addr)$}\\
                         w & \text{otherwise}
                       \end{cases}\\
    \stdUpdatePc{\Phi} &=
                         \begin{cases}
                           \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
                           \begin{array}[t]{l}
                             \text{if $\memreg(\pcreg) = \stdcap$}\\
                             \quad\text{and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}\\
                           \end{array}
                         \end{cases}
  \end{align*}

  \begin{tabular}{|c|p{3.7cm}|p{10.5cm}|}
    \hline
    $i$&$\sem{i}(\Phi)$&Conditions\\
    \hline
    $\fail$&$\failed$&\\
    \hline
    $\halt$&$(\halted,\memheap)$&\\
    \hline
    $\move{\refreg{r_1}}{\rv}$& $\begin{multlined}
      \stdUpdatePc{}(\updateReg{r_1}{\\
          \memreg(\rv)}) \end{multlined}$&\\
    \hline
    $\jmp{\lv}$&$\begin{multlined}
                   \updateReg{\pcreg}{\\\updatePcPerm{\memreg(\lv)}}
                 \end{multlined}$&\\
    \hline
    $\load{\refreg{r_1}}{\refheap{r_2}}$&$\updateReg{r_1}{\memheap(\addr)}$&$\memreg(r_2) = \stdcap$ and $\perm \in \{ \rwx, \rwlx, \exec, \readwrite, \readwritel, \readonly \}$  and  $\start \leq \addr \leq \addrend$\\
    \hline
    $\store{\refheap{r_1}}{\refreg{r_2}}$&$\stdUpdatePc{\updateHeap{\addr}{\var{w}}}$&$\memreg(r_1) = \stdcap$ and $\perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}$  and  $\start \leq \addr \leq \addrend$ and $\var{w} = \memreg(r_2)$
                                                                                       and if  $\var{w} = ((\_,\local),\_,\_,\_)$, then  $\perm \in \{\rwlx,\readwritel \}$\\
    \hline
    $\restricttwo{\refreg{r}}{\rv}$&$\begin{multlined}
                                       \stdUpdatePc{}(\updateReg{r}{\\(\permp',\start,\addrend,\addr)})
                                     \end{multlined}$  &
                                                      $\memreg(r) = \stdcap[\permp]$ and $\permp' = \decodePermPair{\memreg(\rv)}$
                                                      and $\permp'\sqsubseteq \permp$\\
    \hline
    $\geta{\refreg{r_1}}{\refreg{r_2}}$ & $\stdUpdatePc{\updateReg{r_1}{\addr}}$ &
                                                $\memreg(r_2) = ((\_,\_),\_,\_,\addr)$\\
    \hline
    \multicolumn{3}{|c|}{$\cdots$}\\
    \hline
    \_&$\failed$&otherwise\\
    \hline
  \end{tabular}
  \caption{An excerpt from the operational semantics}
  \label{fig:op-sem}
\end{figure*}

@@

The instruction set of our machine is rather basic.  Instructions $i$ include
jumps ($\jmp{}$), conditional jumps ($\jnz{}{}$), a move instruction that copies
a word from one register to another ($\move{}{}$), load and store instructions
for reading from and writing to memory ($\load{}{}$ and $\store{}{}$)

>>>>

@@

The instruction set of our machine is rather basic. Instructions $i$ include
relatively standard jump ($\jmp{}$), conditional jump ($\jnz{}{}$) and move
($\move{}{}$, copies words between registers) instructions. Also familiar are
load and store instructions for reading from and writing to memory ($\load{}{}$
and $\store{}{}$) and an arithmetic addition operator ($\plus{}{}{}$). There are
three instructions for modifying capabilities: $\lea{}{}$ (modifies the current
address), $\restrict{}{}$ (modifies the permission and local/global flag) and
$\subseg{}{}{}$ (modifies the range of a capability). Importantly, these
instructions take care that the resulting capability always carries less
authority than the original (e.g. restrict will only weaken a permission).
Finally, there is an instruction $\isptr{}{}$ for testing whether a word is a
capability or a number and instructions $\getp{}{}$, $\getl{}{}$, $\getb{}{}$,
$\gete{}{}$ and $\geta{}{}$ for obtaining respectively the permissions,
local/global flag, base, end and current address of a capability.

\begin{figure*}
  \centering
  \begin{equation*}
    \Phi  \rightarrow
    \begin{cases}
      \sem{\decode(n)}(\Phi) & \text{if $\memreg(\pcreg) = \stdcap$ and $\start \leq \addr \leq \addrend$ and $\perm \in \{ \exec,\rwx, \rwlx \}$ and $\memheap(\addr) = n$ }\\
      \failed                                 & \text{otherwise}
    \end{cases}
  \end{equation*}\vspace{-.3cm}%
  \begin{align*}
    \updatePcPerm{w} &=
                       \begin{cases}
                         ((\exec,\gl),\start,\addrend,\addr) & \text{if $w = ((\entry,\gl),\start,\addrend,\addr)$}\\
                         w & \text{otherwise}
                       \end{cases}\\
    \stdUpdatePc{\Phi} &=
                         \begin{cases}
                           \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
                           \begin{array}[t]{l}
                             \text{if $\memreg(\pcreg) = \stdcap$}\\
                             \quad\text{and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}\\
                           \end{array}
                         \end{cases}
  \end{align*}

  \begin{tabular}{|c|p{3.8cm}|p{10.4cm}|}
    \hline
    $i$&$\sem{i}(\Phi)$&Conditions\\
    \hline
    $\fail$&$\failed$&\\
    \hline
    $\halt$&$(\halted,\memheap)$&\\
    \hline
    $\move{\refreg{r_1}}{\rv}$& $\begin{multlined}
      \stdUpdatePc{}(\updateReg{r_1}{\\
          \memreg(\rv)}) \end{multlined}$&\\
    \hline
    $\load{\refreg{r_1}}{\refheap{r_2}}$&$\updateReg{r_1}{\memheap(\addr)}$&$\memreg(r_2) = \stdcap$ and $\perm \in \{ \rwx, \rwlx, \exec, \readwrite, \readwritel, \readonly \}$  and  $\start \leq \addr \leq \addrend$\\
    \hline
    $\restricttwo{\refreg{r}}{\rv}$&$\begin{multlined}
                                       \stdUpdatePc{}(\updateReg{r}{\\(\permp',\start,\addrend,\addr)})
                                     \end{multlined}$  &
                                                        $\memreg(r) = \stdcap[\permp]$ and $\permp' = \decodePermPair{\memreg(\rv)}$
                                                      and $\permp'\sqsubseteq \permp$\\
    \hline
    $\geta{\refreg{r_1}}{\refreg{r_2}}$ & $\stdUpdatePc{\updateReg{r_1}{\addr}}$ &
                                                $\memreg(r_2) = ((\_,\_),\_,\_,\addr)$\\
    \hline
    $\jmp{\lv}$&$\begin{multlined}
                   \updateReg{\pcreg}{\\\updatePcPerm{\memreg(\lv)}}
                 \end{multlined}$&\\
    \hline
    $\store{\refheap{r_1}}{\refreg{r_2}}$&$\stdUpdatePc{\updateHeap{\addr}{\var{w}}}$&$\memreg(r_1) = \stdcap$ and $\perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}$  and  $\start \leq \addr \leq \addrend$ and $\var{w} = \memreg(r_2)$
                                                                                       and if  $\var{w} = ((\_,\local),\_,\_,\_)$, then  $\perm \in \{\rwlx,\readwritel \}$\\
    \hline
    \multicolumn{3}{|c|}{$\cdots$}\\
    \hline
    \_&$\failed$&otherwise\\
    \hline
  \end{tabular}
  \caption{An excerpt from the operational semantics}
  \label{fig:op-sem}
\end{figure*}

Figure~\ref{fig:op-sem} shows the operational semantics of our capability
machine, or at least an excerpt containing a number of representative
instructions. Essentially, a configuration $\Phi$ will either decode and execute
the instruction at $\memreg{\pcreg}$ if it is executable and the current address
is in the valid range or fail otherwise. The table in the figure shows for a
given instruction $i$ the result of executing it in configuration $\Phi$. The
$\fail$ and $\halt$ instructions simply fail and halt. The $\move{}{}$
instruction simply modifies the register file as requested and updates the
$\pcreg$ to the next instruction using the meta-function $\stdUpdatePc{}$.

The $\load{}{}$ instruction loads the contents of the requested memory location
into the appropriate register, but only if the capability has appropriate
authority (i.e. read permission and an appropriate range). The $\restrict{}{}$
instruction will update a capability's permissions and global/local status in
the register file, but requires that the new permissions are weaker than the
original and does not turn local capabilities into global ones. The $\geta{}{}$
instruction queries the current address of a capability and stores it in another
register.

The $\jmp{}$ instruction updates the program counter to a requested location,
but it is complicated by the presence of \emph{enter capabilities}, modeled
after the M-Machine's~\cite{Carter:1994:HSF:195473.195579}. Enter capabilities
cannot be used to read, write or execute and their current address and range
cannot be modified. They can only be used to jump to, but when that happens,
their permission is updated to $\exec$. This special behavior makes them
suitable for representing a kind of closures: an opaque package containing a
piece of code together with local encapsulated state. The idea is that one can
build an enter capability $c = ((\entry,\gl),\start,\addrend,\addr)$ where the
range $[\start,\addr-1]$ is used for storing the closure's local state (data or
capabilities) and the range $[\addr,\addrend]$ can be used for storing
instructions. The instructions and data are opaque to an adversary holding $c$
but when $c$ jumped to, the data becomes available to the closure's code through
the updated version of $c$ that is now in the $\pcreg$\footnote{Note that
  $\pcreg$ can be used to $\load{}{}$ from, like any other register.}.

The final semantics we discuss is that of $\store{}{}$. As expected, it updates
the memory to the requested value if the capability has write authority for the
requested location. However, the instruction is complicated by the presence of
\emph{local capabilities}, modeled after the ones in
CHERI~\cite{Watson2015Cheri}. Basically, local capabilities are special in that
they can only be kept in registers, i.e. they cannot be stored to memory. This
means that local capabilities can be \emph{temporarily} given to an adversary,
for the duration of an invocation: if we take care to erase the capability from
the register file after control is passed back to us, we can be sure that he/she
will not have been able to store the capability. However, there is one exception
to the rule above: local capabilities can be stored to a part of memory for
which we have a capability with write-local authority (i.e. permission $\rwl$ or
$\rwlx$). This exception is intended to accomodate regions of memory that behave
like the C stack: stashing away the contents of registers on the stack can be
allowed, even if this includes local capabilities. As long as all capabilities
with write-local authority are themselves local and the stack is cleared after
control is passed back by the adversary, we will see that this does not break
the intended behavior of local capabilities.
